Re-run, Repeat, Reproduce, Reuse, Replicate:\\Transforming Code into Scientific Contributions

### 2017-08-19 draft.tex
Last pass, improving language (WIP).

Replicability is a cornerstone of science.  If an experimental result cannot be
re-obtained by an independent party, it merely becomes, at best, an observation
that may inspire future research (\cite{Mesirov:2010,osc:2015}). Replication
issues have received increased attention in recent years, with a particular
media focus on medicine and psychology (\cite{Iqbal:2016}).  One could think
that computer science would mostly be shielded from such issues, since a
computer program intrinsically describes precisely what it does and is easily
transferred to other researchers.

@@

Have you ever tried to re-run a program you wrote some years ago? Frustratingly, it can often be surprisingly hard. Part of the problem is that technology is evolving at a fast pace and you cannot know in advance how the system, the software and the libraries your program depends on will evolve. Since you wrote the code, you may have reinstalled or upgraded  your operating system. The compiler, interpreter or set of libraries installed may have been replaced with newer versions. You may find yourself battling with arcane issues of library compatibility---thoroughly orthogonal to your immediate research  goals---to execute again a code \emph{that worked perfectly before}. To be clear, it is impossible to write future-proof code, and the best efforts can be stymied by the smallest change in one of the dependencies. At the same time, modernizing an unmaintained ten-year-old code can reveal itself to be an arduous and expensive undertaking, and precarious as that, since each change risks to affect the semantics of the program. Rather trying to predict the future or painstakingly dusting off old code, an often more straightforward and less risky solution is to recreate the old execution environment\footnote{To be clear, and although virtual machines are often a great help here, this is not always possible. It is, however, \emph{always} more difficult when the original execution environment is not documented.}. For this to happen however, the dependencies in terms of systems, software and libraries must be made clear enough.\\

@@

In our case, the R$^0$ version of our tiny walker seems to imply that any version of Python would be fine. This not the case because it uses the print {\em instruction} and the {\tt xrange} operator, both specific to Python 2. The print {\em instruction}, available in Python 2 (a version still widely used; support is scheduled to stop in 2020), has been deprecated in Python 3 (first released in 2008, almost a decade ago) in favor or a  print {\em function}, while the {\tt xrange} operator has been replaced by the {\tt range} operator in Python 3. In order to try to future-proof the code a bit, we might as well target Python 3, as is done in the R$^1$ version. Incidentally, it remains compatible with Python 2. But whichever version is chosen, the crucial step here is to document it.

@@

% Reproducible definition + R^2 not reproducible
The R$^2$ code seems fine enough, but it hides several problems that come to light when trying to {\em reproduce} results. A result is said to be \emph{reproducible} if another researcher can take the original code and input data, execute it, and re-obtain the same result \parencite{Peng:2006}. As explained by \citeauthor{Donoho:2009} \parencite{Donoho:2009}, scientific practice must expect that {\em errors are ubiquitous}, and therefore be robust to them. Ensuring reproducibility is an fundamental step toward this: it provides other researchers the means to verify that your code does indeed produce the published results, and to scrutinize the procedures it uses to produce them. As demonstrated by \citeauthor{Mesnard:2016} \citep{Mesnard:2016}, reproducibility is hard.\\

For instance, the R$^2$ program will not produce the same results all the time. It will, because it is repeatable, produce the same results over repeated executions. But it will not necessarily do so over different execution environments. The cause is to be found in a change that occurred in the pseudo-random number generator between Python 3.2 and Python 3.3. Executed with Python 2.7 to 3.2, the code will produce the sequence 1, 2, 1, 2, 3, 4, 5, 6, 7, 6. But with Python 3.3 to 3.6, it will produce 1, 2, 1, 0, 1, 0, 1, 0, -1, 0.\\

% precise execution environment recording
Because any dependency of a program---to the most basic one, the language itself---can change its behavior from one version to another, executability (R$^1$) and determinism (R$^2$) are necessary but not sufficient for reproducibility. The exact execution environment used to produce the results must also be specified---rather than the broadest set of environments where the code can be effectively run. In other words, assertions such as "the results were obtained with CPython 3.6.1" are more valuable, in a scientific context, than "the program should work with Python 3.x and above". With the increasing complexity of computational stacks, retrieving and deciding what is pertinent (CPU architecture? operating system version? endianness?) might be non-trivial. A good rule of thumb is to include more information than necessary rather than not enough.\\

% if code changes, parameters are lost
Recording the execution environment is not enough though. The R$^2$ program uses a random seed but does not keep a trace of it except in the code. Therefore, should the code change after the production of the results, someone provided with the result file will not be able to know which seed was used, and would need to iterate through all possible random seeds, an impossible task in practice\footnote{Here, with $2^{10}$ possibilities for a 10-step random walk, a seed matching the results could obviously be found. For instance, seed 1123581321 matches the results of R$^0$ with Python 2.7. This is not possible for a 100-step walk.}.

% 1. parameters and 2. code version with results
From this observation, we conclude two things. First, results files should always come alongside their context, i.e. an exhaustive list of the parameters used as well as a precise description of the execution environment, as the R$^3$ code does. Second, the version of the code itself should be recorded as part of the provenance data. It is common for results to have been generated by different versions of the code. Ideally, all results should originate from the same (and last) version of the code. But for long or expensive computations, this may not be feasible. In that case, the result files should contain the version of the code that was used to produce it. This information can be obtained from the version control software. This also allows, if some errors are found and corrected after some results have been obtained, to identify which ones should be recomputed. In R$^3$, the code records the current git hash, and whether the repository holds uncommited changes when the computation starts.\\

@@

% unit test for R^3
Another good practice is to make the code self-verifiable. In R$^3$, a short unit test is provided, that allows the code to verify its own reproducibility. Should this test fail, then there is little hope of reproducing the results. Of course, passing the test is no definitive proof.\\

>>>>

Replicability is a cornerstone of science.  If an experimental result cannot be
re-obtained by an independent party, it merely becomes, at best, an observation
that may inspire future research (\cite{Mesirov:2010,osc:2015}). Replication
issues have received increased attention in recent years, with a particular focus on medicine and psychology (\cite{Iqbal:2016}).  One could think
that computer science would mostly be shielded from such issues, since a
computer program intrinsically describes precisely what it does and is easily
disseminated to other researchers without alteration.

@@

Have you ever tried to re-run a program you wrote some years ago? Frustratingly, it can often be surprisingly hard. Part of the problem is that technology is evolving at a fast pace and you cannot know in advance how the system, the software and the libraries your program depends on will evolve. Since you wrote the code, you may have reinstalled or upgraded  your operating system. The compiler, interpreter or set of libraries installed may have been replaced with newer versions. You may find yourself battling with arcane issues of library compatibility---thoroughly orthogonal to your immediate research  goals---to execute again a code \emph{that worked perfectly before}. To be clear, it is impossible to write future-proof code, and the best efforts can be stymied by the smallest change in one of the dependencies. At the same time, modernizing an unmaintained ten-year-old code can reveal itself to be an arduous and expensive undertaking, and precarious as that, since each change risks affecting the semantics of the program. Rather trying to predict the future or painstakingly dusting off old code, an often more straightforward and less risky solution is to recreate the old execution environment\footnote{To be clear, and although virtual machines are often a great help here, this is not always possible. It is, however, \emph{always} more difficult when the original execution environment is not documented.}. For this to happen however, the dependencies in terms of systems, software and libraries must be made clear enough.\\

@@

In our case, the R$^0$ version of our tiny walker seems to imply that any version of Python would be fine. This not the case: it uses the print {\em instruction} and the {\tt xrange} operator, both specific to Python 2. The print {\em instruction}, available in Python 2 (a version still widely used; support is scheduled to stop in 2020), has been deprecated in Python 3 (first released in 2008, almost a decade ago) in favor or a  print {\em function}, while the {\tt xrange} operator has been replaced by the {\tt range} operator in Python 3. In order to try to future-proof the code a bit, we might as well target Python 3, as is done in the R$^1$ version. Incidentally, it remains compatible with Python 2. But whichever version is chosen, the crucial step here is to document it.

@@

% Reproducible definition + R^2 not reproducible
The R$^2$ code seems fine enough, but it hides several problems that come to light when trying to {\em reproduce} results. A result is said to be \emph{reproducible} if another researcher can take the original code and input data, execute it, and re-obtain the same result \parencite{Peng:2006}. As explained by \citeauthor{Donoho:2009} \parencite{Donoho:2009}, scientific practice must expect that {\em errors are ubiquitous}, and therefore be robust to them. Ensuring reproducibility is an fundamental step toward this: it provides other researchers the means to verify that your code does indeed produce the published results, and to scrutinize the procedures it used to produce them. As demonstrated by \citeauthor{Mesnard:2016} \citep{Mesnard:2016}, reproducibility is hard.\\

For instance, the R$^2$ program will not produce the same results all the time. It will, because it is repeatable, produce the same results over repeated executions. But it will not necessarily do so over different execution environments. The cause is to be found in a change that occurred in the pseudo-random number generator between Python 3.2 and Python 3.3. Executed with Python 2.7 to 3.2, the code will produce the sequence 1, 2, 1, 2, 3, 4, 5, 6, 7, 6. But with Python 3.3 to 3.6, it will produce 1, 2, 1, 0, 1, 0, 1, 0, -1, 0. With future version of the language, it may change still.\\

% precise execution environment recording
Because any dependency of a program---to the most basic one, the language itself---can change its behavior from one version to another, executability (R$^1$) and determinism (R$^2$) are necessary but not sufficient for reproducibility. The exact execution environment used to produce the results must also be specified---rather than the broadest set of environments where the code can be effectively run. In other words, assertions such as "the results were obtained with CPython 3.6.1" are more valuable, in a scientific context, than "the program works with Python 3.x and above". With the increasing complexity of computational stacks, retrieving and deciding what is pertinent (CPU architecture? operating system version? endianness?) might be non-trivial. A good rule of thumb is to include more information than necessary rather than not enough, and some rather than none.\\

% if code changes, parameters are lost
Recording the execution environment is not enough though. The R$^2$ program uses a random seed but does not keep a trace of it except in the code. Should the code change after the production of the results, someone provided with the last version of the code will not be able to know which seed was used to produce the results, and would need to iterate through all possible random seeds, an impossible task in practice\footnote{Here, with $2^{10}$ possibilities for a 10-step random walk, a seed matching the results could obviously be found. For instance, seed 1123581321 matches the results of R$^0$ with Python 2.7. This is not possible for a 100-step walk.}.\\

% 1. parameters and 2. code version with results
This is why results files should come alongside their context, i.e. an exhaustive list of the parameters used as well as a precise description of the execution environment, as the R$^3$ code does. The code itself is part of that context: the version of the code must be recorded as part of the provenance data. It is common for different results or different figures to have been generated by different versions of the code. Ideally, all results should originate from the same (and last) version of the code. But for long or expensive computations, this may not be feasible. In that case, the result files should contain the version of the code that was used to produce it. This information can be obtained from the version control software. This also allows, if some errors are found and corrected after some results have been obtained, to identify which ones should be recomputed. In R$^3$, the code records the current git hash, and whether the repository holds uncommited changes when the computation starts.\\

@@

% unit test for R^3
Another good practice is to make the code self-verifiable. In R$^3$, a short unit test is provided, that allows the code to verify its own reproducibility. Should this test fail, then there is little hope of reproducing the results. Of course, passing the test does not guarantee anything.\\
